### 2018-05-16 submission/manuscript.Rmd
re-write results section

***The Number of OTUs are Dependent on Polymerase:***
A consistent difference in the range of number of OTUs, that was dependent on the polymerase used was observed for fecal samples [Figure 1]. Additionally, there was a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) between polymerases [Figure 1]. For fecal samples, there was a significant difference between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1]. No differences between the specific groups was found based on a Tukey post-hoc test (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2]. This polymerase dependent difference in the range of number of OTUs also was observed in the mock community samples [Figure 2]. Regardless if fecal or mock communities were used, the same polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S4]. In contrast to the results obtained with fecal samples, differences between polymerases were observed as early as 20 cycles in the mock community samples (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3]. For the mock community, the majority of  differences in the number of OTUs were between Kappa and Platinum and the other polymerases across different number of cycles [Table S4]. Based on these observations in fecal and mock communities, it is clear that different polymerases result in a different total number of OTUs within a sample.

***Minimal Bray-Curtis Differences are Detected and are Dependent on Cycle Number:***
A few small differences based on the number of cycles were detected in overall bacterial community composition. Within the same fecal sample and independent of polymerases, there were large differences in the community composition between 20 and 25 cycles [Figure 3A]. Additionally, when data was available for the mock communities, there were also large observed differences between 20 and 25 cycles [Figure 3B]. These community differences within a sample disappear when comparing 25 to 30 cycles and do not persist past 25 cycles [Figure 3]. Although these trends are clearly noticeable, we found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05). Using PERMANOVA to test for community differences based on the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`. For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 at 15 cycles. Overall, these data suggest that there are small polymerase differences in Bray-Curtis index that are dependent on the number of cycles.

***Sequence Error is Dependent on both Polymerase and Cycle Number:***
Differences in the median error rate varied by polymerase across the number of cycles used [Table S5]. The highest error rates were for the Kappa HiFi DNA polymerase [Figure 4]. The differences in the median error rate between the different polymerases was dependent on the number of cycles with Platinum and Kappa having the most differences versus other polymerases [Figure 4 and Table S6]. The total sequences with at least one error was also cycle number dependent [Table S7 & S8. These differences in sequences with at least one error were mostly due to differences in Accuprime^TM^ and Kappa versus the other polymerases [Table S7 & S8]. Finally, we did not observe a polymerase dependent difference on base substitution rate [Figure S1]. These results suggest that sequence error is dependent on polymerase. We next wanted to assess whether chimeras also were dependent on polymerase and whether this affected OTU number.

***Prevalence of Chimeric Sequences are Polymerase Dependent and Correlate with the Number of OTUs:***
There were significant differences in the chimera prevalence based on polymerase used at all numbers of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9]. Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S10]. Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B]. There was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C]. A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 6]. This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 6]. This data suggests that chimera prevalence depends on polymerase used and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.

@@

In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase used [Figure 1-2 & 4-5]. These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g. richness). Additionally, the earlier detection of differences in total number of OTUs between polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples. We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences. This suggests that specific diversity differences between studies can be attributed to differences in polyermase used. Based on our observations metrics that measure within sample diversity depend on polymerase but this may not be the case for metrics that assess between sample diversity.

There were few differences that depend on polymerase for between sample diversity, as measured by the Bray-Curtis index. Using this metric our observations generally found no differences in the overall bacterial community composition for the number of cycles used. One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group. Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed polymerase dependent differences in chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in polymerases. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. With differences in the number of OTUs and chimera prevalence depending on polymerase used, it might be easier to avoid specific DNA polymerase families altogether.

Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, this is unlikely to be the only contributor. For example, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the material safety data sheet (MSDS) the differences between the two polymerases are not immediately apparent. Both polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the polymerases. We are unlikely to avoid adding polymerase dependent bias to 16S rRNA gene sequencing results, however, these differences may not be large enough to mask the actual biological signal.

The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence that can be masked by biological differences. The sequence error introduced by the polymerase is also small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. The choice of polymerase should be an important consideration in the creation of a 16S rRNA gene sequencing workflow because using different polymerases can add error and bias to the downstream observations. Although standardization of the workflow may partially solve this problem by introducing a consistent bias to all samples, it does come at a cost.

>>>>

***The Number of OTUs are Dependent on Polymerase:***
Differences in the range of the number of OTUs detected for fecal samples is dependent on the polymerase used [Figure 1]. Additionally, there is a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases [Figure 1]. There is an overall difference in the number of OTUs detected within fecal samples between polymerases at 35 cycles (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S1]. No differences between the polymerases was found based on a Tukey post-hoc test (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S2]. The polymerase dependent difference in the range of the number of OTUs also was observed in the mock community samples [Figure 2]. Regardless if fecal or mock communities were used, the same polymerases had high (Platinum) and low (Accuprime) number of OTUs and this was consistent across the number of cycles used [Figure 1-2 & Table S1-S4]. In contrast to the results obtained with fecal samples, differences between polymerases were observed as early as 20 cycles in the mock community (F-stat = `r round(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "F.value"], digits = 2)`, P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`) [Table S3]. The majority of differences for the number of OTUs detected in the Mock community is largely due to Kappa and Platinum differences to the other polymerases across the different number of cycles [Table S4]. Based on these observations in fecal and mock communities, it is clear that different polymerases result in a different total number of OTUs within a sample.

***Minimal Bray-Curtis Differences are Detected and are Dependent on Cycle Number:***
A few small differences based on the number of cycles were detected in overall bacterial community composition. Within the same fecal sample and independent of polymerases, there are large differences in the community composition between 20 and 25 cycles [Figure 3A]. Additionally, when data was available for the mock communities, there are also large observed differences between 20 and 25 cycles [Figure 3B]. These community differences within a sample disappear when comparing 25 to 30 cycles and do not persist past 25 cycles [Figure 3]. Although these trends are clearly noticeable, there is no detectable difference in Bray-Curtis index when comparing to the previous 5-cycle increment for both fecal and mock communities after multiple comparison correction (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05). Using PERMANOVA to test for community differences based on the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`. For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 at 15 cycles. Overall, these data suggest that there are small polymerase differences in Bray-Curtis index that are dependent on the number of cycles.

***Sequence Error is Dependent on both Polymerase and Cycle Number:***
Differences in the median error rate varied by polymerase, with Kappa being the highest [Figure 4 & Table S5]. The differences in the median error rate between the polymerases also is dependent on the number of cycles, with Platinum and Kappa having the most differences versus other polymerases [Figure 4 and Table S6]. The total sequences with at least one error also is dependent on the number of cycles [Table S7 & S8]. These differences in sequences with at least one error is mostly due to differences in Accuprime^TM^ and Kappa versus the other polymerases [Table S7 & S8]. The differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. These results suggest that sequence error is dependent on polymerase and the number of cycles, but is not attributable to a specific type. Based on these results, we examined whether chimeras also were dependent on polymerase and whether this also could affect the number of OTUs.

***Prevalence of Chimeric Sequences are Polymerase Dependent and Correlate with the Number of OTUs:***
There is significant differences in the chimera prevalence based on polymerase at all the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S9]. Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S10]. Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 5A & 5B]. Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 5C]. A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 6]. This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 6]. This data suggests that chimera prevalence depends on polymerase used and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.

@@

In this study we show that the number of OTUs, error rate, and chimera prevalence depends on polymerase used [Figure 1-5]. These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g. richness). Additionally, the earlier detection of differences in total number of OTUs between polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples. We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences. This suggests that some of the specific diversity differences between studies can be attributed to differences in polyermase used. Based on these observations, metrics that measure within sample diversity like richness depend on polymerase but this may not be the case for metrics that assess between sample diversity.

There were few differences that depend on polymerase for between sample diversity, as measured by the Bray-Curtis index. Our observations generally found no differences in the overall bacterial community composition for the number of cycles used. One possible reason for this outcome is that our study did not have enough power to detect differences due to low sample number in each group. Another reason is that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed polymerase dependent differences in chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in polymerases. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. With differences in the number of OTUs and chimera prevalence depending on polymerase used, it might be easier to avoid specific DNA polymerase families altogether.

Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, this is unlikely to be the only contributor. For example, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the material safety data sheet the differences between the two polymerases are not immediately apparent. Both polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the polymerases. We are unlikely to avoid adding polymerase dependent bias to 16S rRNA gene sequencing results, however, these differences may not be large enough to mask the actual biological signal.

The majority of polymerases we studied add small increases in the number of OTUs and chimera prevalence and may be masked by biological differences. The sequence error introduced by the polymerase is also small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. The choice of polymerase should be an important consideration in the creation of a 16S rRNA gene sequencing workflow because using different polymerases can add error and bias to the downstream observations. Although standardization of the workflow may partially solve this problem by introducing a consistent bias to all samples, it does come at a cost.

### 2018-05-16 submission/manuscript.tex
re-write results section

\textbf{\emph{The Number of OTUs are Dependent on Polymerase:}} A
consistent difference in the range of number of OTUs, that was dependent
on the polymerase used was observed for fecal samples {[}Figure 1{]}.
Additionally, there was a trend for lower number of cycles (15-20) to
result in a reduced range in the number of OTUs versus higher number of
cycles (25, 30, and 35) between polymerases {[}Figure 1{]}. For fecal
samples, there was a significant difference between polymerases at 35
cycles (F-stat \textgreater{} 16.35, P-value = 9.7e-05) {[}Table S1{]}.
No differences between the specific groups was found based on a Tukey
post-hoc test (P-value \textgreater{} 0.05) {[}Table S2{]}. This
polymerase dependent difference in the range of number of OTUs also was
observed in the mock community samples {[}Figure 2{]}. Regardless if
fecal or mock communities were used, the same polymerases had high
(Platinum) and low (Accuprime) number of OTUs and this was consistent
across the number of cycles used {[}Figure 1-2 \& Table S1-S4{]}. In
contrast to the results obtained with fecal samples, differences between
polymerases were observed as early as 20 cycles in the mock community
samples (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. For the mock
community, the majority of differences in the number of OTUs were
between Kappa and Platinum and the other polymerases across different
number of cycles {[}Table S4{]}. Based on these observations in fecal
and mock communities, it is clear that different polymerases result in a
different total number of OTUs within a sample.

\textbf{\emph{Minimal Bray-Curtis Differences are Detected and are
Dependent on Cycle Number:}} A few small differences based on the number
of cycles were detected in overall bacterial community composition.
Within the same fecal sample and independent of polymerases, there were
large differences in the community composition between 20 and 25 cycles
{[}Figure 3A{]}. Additionally, when data was available for the mock
communities, there were also large observed differences between 20 and
25 cycles {[}Figure 3B{]}. These community differences within a sample
disappear when comparing 25 to 30 cycles and do not persist past 25
cycles {[}Figure 3{]}. Although these trends are clearly noticeable, we
found that there was no detectable difference in Bray-Curtis index when
comparing to the previous 5-cycle increment for both fecal and mock
communities after multiple comparison correction (P-value \textgreater{}
0.05). Using PERMANOVA to test for community differences based on the
number of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03. For fecal samples, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 at 15
cycles. Overall, these data suggest that there are small polymerase
differences in Bray-Curtis index that are dependent on the number of
cycles.

\textbf{\emph{Sequence Error is Dependent on both Polymerase and Cycle
Number:}} Differences in the median error rate varied by polymerase
across the number of cycles used {[}Table S5{]}. The highest error rates
were for the Kappa HiFi DNA polymerase {[}Figure 4{]}. The differences
in the median error rate between the different polymerases was dependent
on the number of cycles with Platinum and Kappa having the most
differences versus other polymerases {[}Figure 4 and Table S6{]}. The
total sequences with at least one error was also cycle number dependent
{[}Table S7 \& S8. These differences in sequences with at least one
error were mostly due to differences in Accuprime\textsuperscript{TM}
and Kappa versus the other polymerases {[}Table S7 \& S8{]}. Finally, we
did not observe a polymerase dependent difference on base substitution
rate {[}Figure S1{]}. These results suggest that sequence error is
dependent on polymerase. We next wanted to assess whether chimeras also
were dependent on polymerase and whether this affected OTU number.

\textbf{\emph{Prevalence of Chimeric Sequences are Polymerase Dependent
and Correlate with the Number of OTUs:}} There were significant
differences in the chimera prevalence based on polymerase used at all
numbers of cycles used (P-value \textless{} 0.05) {[}Table S9{]}.
Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S10{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 5A \& 5B{]}. There was a plateau in the total percent
of chimeras that were removed that was similar for all polymerases
{[}Figure 5C{]}. A positive correlation was observed between chimeric
sequences and the number of OTUs for all polymerases {[}Figure 6{]}.
This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
This data suggests that chimera prevalence depends on polymerase used
and confirms that the number of OTUs is dependent on the prevalence of
these chimeric sequences.

@@

In this study we show that the number of OTUs, error rate, and chimera
prevalence depends on polymerase used {[}Figure 1-2 \& 4-5{]}. These
differences are important because many diversity metrics rely on the
number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between polymerases in the mock versus fecal samples might indicate
that high biomass samples may underestimate the biases present within
low biomass samples. We observed that undetected chimeras that were not
identified and removed using standard bioinformatic approaches cause
many of these differences. This suggests that specific diversity
differences between studies can be attributed to differences in
polyermase used. Based on our observations metrics that measure within
sample diversity depend on polymerase but this may not be the case for
metrics that assess between sample diversity.

There were few differences that depend on polymerase for between sample
diversity, as measured by the Bray-Curtis index. Using this metric our
observations generally found no differences in the overall bacterial
community composition for the number of cycles used. One possible reason
for this outcome was that our study did not have enough power to detect
differences due to low sample number in each group. Another reason was
that many of the OTUs are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight chimeric OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate these observed polymerase
dependent differences in chimera prevalence. Metrics that solely use
presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996)) may be
less robust to chimera prevalence and by extension total number of OTU
differences in polymerases. When choosing a distance metric, careful
consideration of the biases introduced from the PCR step of the 16S rRNA
gene sequencing workflow need to be taken into account. With differences
in the number of OTUs and chimera prevalence depending on polymerase
used, it might be easier to avoid specific DNA polymerase families
altogether.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family, this is unlikely to be the only
contributor. For example, the highest and lowest chimera rates both
belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the material safety data sheet (MSDS) the
differences between the two polymerases are not immediately apparent.
Both polymerases contain a recombinant \emph{Taq} DNA polymerase, a
\emph{Pyrococcus} spp GB-D polymerase and a platinum \emph{Taq}
antibody. With everything else being equal, it is possible that
differences in how the recombinant \emph{Taq} was generated could be a
contributing factor for the differences observed between the
polymerases. We are unlikely to avoid adding polymerase dependent bias
to 16S rRNA gene sequencing results, however, these differences may not
be large enough to mask the actual biological signal.

The majority of polymerases we studied add small increases in the number
of OTUs and chimera prevalence that can be masked by biological
differences. The sequence error introduced by the polymerase is also
small and likely to be smaller than the biological variation within a
specific study, which would be consistent with previous findings for
preservation and DNA extraction methods (Salter et al., 2014; Song et
al., 2016; Luo et al., 2016). The choice of polymerase should be an
important consideration in the creation of a 16S rRNA gene sequencing
workflow because using different polymerases can add error and bias to
the downstream observations. Although standardization of the workflow
may partially solve this problem by introducing a consistent bias to all
samples, it does come at a cost.

>>>>

\textbf{\emph{The Number of OTUs are Dependent on Polymerase:}}
Differences in the range of the number of OTUs detected for fecal
samples is dependent on the polymerase used {[}Figure 1{]}.
Additionally, there is a trend for lower number of cycles (15-20) to
result in a reduced range in the number of OTUs versus higher number of
cycles (25, 30, and 35) for all polymerases {[}Figure 1{]}. There is an
overall difference in the number of OTUs detected within fecal samples
between polymerases at 35 cycles (F-stat \textgreater{} 16.35, P-value =
9.7e-05) {[}Table S1{]}. No differences between the polymerases was
found based on a Tukey post-hoc test (P-value \textgreater{} 0.05)
{[}Table S2{]}. The polymerase dependent difference in the range of the
number of OTUs also was observed in the mock community samples {[}Figure
2{]}. Regardless if fecal or mock communities were used, the same
polymerases had high (Platinum) and low (Accuprime) number of OTUs and
this was consistent across the number of cycles used {[}Figure 1-2 \&
Table S1-S4{]}. In contrast to the results obtained with fecal samples,
differences between polymerases were observed as early as 20 cycles in
the mock community (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. The
majority of differences for the number of OTUs detected in the Mock
community is largely due to Kappa and Platinum differences to the other
polymerases across the different number of cycles {[}Table S4{]}. Based
on these observations in fecal and mock communities, it is clear that
different polymerases result in a different total number of OTUs within
a sample.

\textbf{\emph{Minimal Bray-Curtis Differences are Detected and are
Dependent on Cycle Number:}} A few small differences based on the number
of cycles were detected in overall bacterial community composition.
Within the same fecal sample and independent of polymerases, there are
large differences in the community composition between 20 and 25 cycles
{[}Figure 3A{]}. Additionally, when data was available for the mock
communities, there are also large observed differences between 20 and 25
cycles {[}Figure 3B{]}. These community differences within a sample
disappear when comparing 25 to 30 cycles and do not persist past 25
cycles {[}Figure 3{]}. Although these trends are clearly noticeable,
there is no detectable difference in Bray-Curtis index when comparing to
the previous 5-cycle increment for both fecal and mock communities after
multiple comparison correction (P-value \textgreater{} 0.05). Using
PERMANOVA to test for community differences based on the number of
cycles within polymerases, only Phusion had cycle dependent differences
(P-value = 0.03. For fecal samples, Phusion was one of two polymerases
that had enough sequences to be rarefied to 1000 at 15 cycles. Overall,
these data suggest that there are small polymerase differences in
Bray-Curtis index that are dependent on the number of cycles.

\textbf{\emph{Sequence Error is Dependent on both Polymerase and Cycle
Number:}} Differences in the median error rate varied by polymerase,
with Kappa being the highest {[}Figure 4 \& Table S5{]}. The differences
in the median error rate between the polymerases also is dependent on
the number of cycles, with Platinum and Kappa having the most
differences versus other polymerases {[}Figure 4 and Table S6{]}. The
total sequences with at least one error also is dependent on the number
of cycles {[}Table S7 \& S8{]}. These differences in sequences with at
least one error is mostly due to differences in
Accuprime\textsuperscript{TM} and Kappa versus the other polymerases
{[}Table S7 \& S8{]}. The differences in error rates were not due to
polymerase dependent differences in base substitution rate {[}Figure
S1{]}. These results suggest that sequence error is dependent on
polymerase and the number of cycles, but is not attributable to a
specific type. Based on these results, we examined whether chimeras also
were dependent on polymerase and whether this also could affect the
number of OTUs.

\textbf{\emph{Prevalence of Chimeric Sequences are Polymerase Dependent
and Correlate with the Number of OTUs:}} There is significant
differences in the chimera prevalence based on polymerase at all the
number of cycles used (P-value \textless{} 0.05) {[}Table S9{]}.
Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S10{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 5A \& 5B{]}. Additionally, there was a plateau in the
total percent of chimeras that were removed that was similar for all
polymerases {[}Figure 5C{]}. A positive correlation was observed between
chimeric sequences and the number of OTUs for all polymerases {[}Figure
6{]}. This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
This data suggests that chimera prevalence depends on polymerase used
and confirms that the number of OTUs is dependent on the prevalence of
these chimeric sequences.

@@

In this study we show that the number of OTUs, error rate, and chimera
prevalence depends on polymerase used {[}Figure 1-5{]}. These
differences are important because many diversity metrics rely on the
number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between polymerases in the mock versus fecal samples might indicate
that high biomass samples may underestimate the biases present within
low biomass samples. We observed that undetected chimeras that were not
identified and removed using standard bioinformatic approaches cause
many of these differences. This suggests that some of the specific
diversity differences between studies can be attributed to differences
in polyermase used. Based on these observations, metrics that measure
within sample diversity like richness depend on polymerase but this may
not be the case for metrics that assess between sample diversity.

There were few differences that depend on polymerase for between sample
diversity, as measured by the Bray-Curtis index. Our observations
generally found no differences in the overall bacterial community
composition for the number of cycles used. One possible reason for this
outcome is that our study did not have enough power to detect
differences due to low sample number in each group. Another reason is
that many of the OTUs are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight chimeric OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate these observed polymerase
dependent differences in chimera prevalence. Metrics that solely use
presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996)) may be
less robust to chimera prevalence and by extension total number of OTU
differences in polymerases. When choosing a distance metric, careful
consideration of the biases introduced from the PCR step of the 16S rRNA
gene sequencing workflow need to be taken into account. With differences
in the number of OTUs and chimera prevalence depending on polymerase
used, it might be easier to avoid specific DNA polymerase families
altogether.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family, this is unlikely to be the only
contributor. For example, the highest and lowest chimera rates both
belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the material safety data sheet the differences
between the two polymerases are not immediately apparent. Both
polymerases contain a recombinant \emph{Taq} DNA polymerase, a
\emph{Pyrococcus} spp GB-D polymerase and a platinum \emph{Taq}
antibody. With everything else being equal, it is possible that
differences in how the recombinant \emph{Taq} was generated could be a
contributing factor for the differences observed between the
polymerases. We are unlikely to avoid adding polymerase dependent bias
to 16S rRNA gene sequencing results, however, these differences may not
be large enough to mask the actual biological signal.

The majority of polymerases we studied add small increases in the number
of OTUs and chimera prevalence and may be masked by biological
differences. The sequence error introduced by the polymerase is also
small and likely to be smaller than the biological variation within a
specific study, which would be consistent with previous findings for
preservation and DNA extraction methods (Salter et al., 2014; Song et
al., 2016; Luo et al., 2016). The choice of polymerase should be an
important consideration in the creation of a 16S rRNA gene sequencing
workflow because using different polymerases can add error and bias to
the downstream observations. Although standardization of the workflow
may partially solve this problem by introducing a consistent bias to all
samples, it does come at a cost.
